---
layout: post
title: "Selecting a matched subsample"
author: "Matthew Peverill"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: gfm
    preserve_yaml: true
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile,
                      encoding = encoding,
                      output_file = file.path(paste0(
                                                  "~/Dropbox/mrpeverill-website/_posts/",
                                                  Sys.Date(),
                                                  '-',
                                                  substr(basename(inputFile), 1, nchar(basename(inputFile)) - 4),
                                                  '.md'
                                                  )
                                              )
                      )
    })
---

```{r setup, include=FALSE,cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(base.dir = "~/Dropbox/mrpeverill-website/", base.url = "/")
knitr::opts_chunk$set(fig.path = "assets/img/MatchedSamplingTest/")
```

This is a second post in a series on splitting samples. In this case, say you have a very small sub-group of a large sample. You want to look at that subgroup and controls, but you don't want your sample to be 90% controls. Instead, you want the subgroup and a sub-sample of controls matched on some demographic variables. As a further complication, lets make one variable (age) continuous, and lets make age and sex correlated with subgroup membership. This example is heavily cribbed from a [post by Norbert Köhler](https://datascienceplus.com/how-to-use-r-for-matching-samples-propensity-score/).


```{r,warning=FALSE,results=FALSE,message=FALSE}
library(sn)
library(tidyverse); library(ggthemes); theme_set(theme_tufte())
library(ggExtra)
library(pander)
library(MatchIt)
library(simstudy)
```

# Simulation

```{r}
set.seed(31453)

simdef<-defData(varname="age",
                dist="uniformInt",
                formula="120;144") #age in months between 10-12
simdef<-defData(simdef,varname="sex",
                dist="binary",
                formula=".5")
simdef<-defData(simdef,varname="parent.ed",
                dist="categorical",
                formula=genCatFormula(n=6))
simdef<-defData(simdef,varname="missingdata",
                dist="binary",
                formula=".2")
simdef<-defData(simdef,varname="inSubGroup",
                dist="binary",
                formula=".005/12 * (age-132) + .005*sex + .0175")


df<-genData(12000,simdef)

df$income<-rsn(nrow(df),alpha=3)
numbers_of_bins = 10
df <- df %>%
    mutate(
        # bin i:
        i.bin = cut(income,
                    breaks = unique(quantile(
                        income,
                        probs = seq.int(0, 1, by = 1 / numbers_of_bins)
                    )),
                    include.lowest = TRUE,
                    labels=FALSE)
    )

df<-as.data.frame(df)

factorialize<-c("sex","missingdata","parent.ed","inSubGroup")
df[factorialize] <- lapply(df[factorialize], factor)
levels(df$inSubGroup)<-c("control","treatment")

pander(head(df))

pander(table(df$inSubGroup))
```

# Is there an imbalance?

We only need to match on age, sex, and one other categorical variable.

```{r}
imbalance_model <-
    matchit(
        inSubGroup ~ age +sex + i.bin,
        data = df,
        method = NULL,
        distance = "glm"
    )

summary(imbalance_model)
```
Yes, age and sex are imbalanced (which we simulated). So is income!

# Nearest Neighbor Matching

Note that it is important to code variable type correctly, i.e. that factors are factors and not numeric.

## Sub-sampling

We want 2 controls per 'treatment' participant.

```{r}
matching_model <-
    matchit(
        inSubGroup ~ age + sex + i.bin,
        data = df,
        method = "nearest",
        distance = "glm",
        ratio= 2
    )

summary(matching_model,un=FALSE)
```

The distribution parameters should be similar, and the control n should be twice the treated n. Then we save the new data frame:

```{r}
df.match<-match.data(matching_model)[,1:ncol(df)]
```

## Checking


```{r}
df.match$inSubGroup<-factor(df.match$inSubGroup,labels=c("Control","Treatment"))
histbygroup<-function(plotdata,xvar="i") {
	ggplot(plotdata,aes_string(x=xvar)) +
		geom_density() +
		facet_grid(~inSubGroup)
}

histbygroup(df.match,"i.bin")
histbygroup(df.match,"age")

barbygroup<-function(plotdata,xvar="i") {
	ggplot(plotdata,aes_string(x=xvar)) +
		geom_bar() +
		facet_grid(~inSubGroup)
}

barbygroup(df.match,"sex")
```

Nearest neighbor is doing pretty well! Here's another method for comparison:

# Optimal Matching

## Sub-sampling

We want 2 controls per 'treatment' participant.

```{r}
matching_model2 <-
    matchit(
        inSubGroup ~ age + sex + i.bin,
        data = df,
        method = "optimal",
        distance = "glm",
        ratio= 2
    )

summary(matching_model2,un=FALSE)
```

The distribution parameters should be similar, and the control n should be twice the treated n. Then we save the new data frame:

```{r}
df.match2<-match.data(matching_model2)[,1:ncol(df)]
```

## Checking


```{r}
histbygroup<-function(plotdata,xvar="i") {
	ggplot(plotdata,aes_string(x=xvar)) +
		geom_density() +
		facet_grid(~inSubGroup)
}

histbygroup(df.match2,"i.bin")
histbygroup(df.match2,"age")

barbygroup<-function(plotdata,xvar="i") {
	ggplot(plotdata,aes_string(x=xvar)) +
		geom_bar() +
		facet_grid(~inSubGroup)
}

barbygroup(df.match2,"sex")
```

Nearest neighbor appears to do a better job
